### 2016-10-04 Paper_v7.tex

\begin{titlepage}
{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\vspace{0.05in}
{\par \centering Russell Gould$^{\ddag}$ and David M. Bassen, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
\vspace{0.05in}
{\par \centering Robert Frederick Smith School of Chemical and Biomolecular Engineering and $^{\ddag}$Nancy E. and Peter C. Meinig School of Biomedical Engineering, Cornell University, Ithaca NY 14853}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
\vspace{0.1in}
% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
%\vspace{0.1in}
{\par \centering \textbf{Character count:}~\emph{56,659}}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
\vspace{0.1in}
{\par \centering Jonathan T. Butcher,}
{\par \centering Associate Professor, Nancy E. and Peter C. Meinig School of Biomedical Engineering,}
{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jtb47@cornell.edu}
{\par \centering Phone: (607) 255 - 3575}
{\par \centering Fax: (607) 255 - 7330}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
We used ordinary differential equations (ODEs) to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was enhanced following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

\subsection*{Formulation, solution and analysis of the EMT model equations}
\subsubsection*{EMT signaling events.}
EMT signaling events were modeled using either saturation or mass-action kinetics within an ordinary differential equation (ODE) framework:
\begin{equation}
	\frac{1}{\tau_{i}}\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)-\mu x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}\\
\end{equation}
where $\mathcal{R}$ denotes the number of signaling reactions and $\mathcal{M}$ denotes the number of signaling proteins in the model.
The quantity $\tau_{i}$ denotes a time scale parameter for species $i$ which captures un-modeled effects; in the current study $\tau_{i} = 1$ for all species.
The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$.
Typically, reaction $j$ is a non-linear function of biochemical and enzyme species abundance, as well as unknown model parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$.
If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$.
Conversely, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
Species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.

@@

\subsubsection*{EMT gene expression processes.}
The EMT model described both signal transduction and gene expression events following the addition of TGF$-\beta$ and VEGF-A.
For each gene of the $\mathcal{G}=23$ we considered, we modeled both the resulting mRNA ($m_{j}$) and protein ($p_{j}$):
\begin{eqnarray}
	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\delta_{m,j}\right)m_{j}+\lambda_{j}\\
	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\delta_{p,j}\right)p_{j}
\end{eqnarray}where $j=1,2,\dots,\mathcal{G}$.
The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
the terms $\delta_{m,j}$ and $\delta_{p,j}$ denote degradation constants for mRNA and protein, respectively.
Lastly, $\mu$ denotes the specific growth rate, and $\lambda_{j}$ denotes the constitutive rate of gene expression for gene $j$.
The specific transcription rate was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
abundance of transcription factors, or other regulators influenced the expression of gene $j$.
The kinetic rate of transcription was modeled as:
\begin{equation}
	\bar{r}_{T,j} = \alpha_{j}\left[V^{max}_{T}\left(\frac{G_{j}}{K_{T}+G_{j}}\right)\right]
\end{equation}where the maximum gene expression rate was defined as the product of a characteristic transcription rate constant ($k_{T}$)
and the abundance of RNA polymerase, $V^{max}_{T} = k_{T}\left(RNAP\right)$. The parameter $\alpha_{j}$ denotes the gene specific correction to the characteristic transcription rate.
(estimated in this study), while $k_{T}$, $G_{j}$ and $RNAP$ were estimated from literature \citep{Milo:2010aa}.
Similar to the signaling processes, the gene expression control term $0\leq u_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
For each rate, we used a rule-based approach to select from competing control factors.
If the expression of gene $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$u_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on the expression of gene $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \citep{pr3010178,pr3010138}.
If a gene expression process has no modifying factors, $u_{j}=1$.
Lastly, the specific translation rate was modeled as:
\begin{equation}
	r_{X,j} = \beta_{j}\left[V^{max}_{X}\left(\frac{m_{j}}{K_{X}+m_{j}}\right)\right]
\end{equation}where $V^{max}_{X}$ denotes a characteristic maximum translation rate estimated from literature, $\beta_{j}$ denotes the transcript specific correction the characteristic translation rate,
and $K_{X}$ denotes a translation saturation constant.
The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
and the abundance of Ribosomes ($RIBO$), $V^{max}_{X} = k_{X}\left(RIBO\right)$, where both $k_{X}$ and $RIBO$ abundance were estimated from literature \citep{Milo:2010aa}.

@@

\subsubsection*{Estimation of model parameters using multiobjective optimization.}
The EMT model had 296 unknown parameters (169 kinetic constants, 44 saturation constants, 38 control logic parameters, and 45 non-zero initial conditions) which were not uniquely identifiable given the training data.
Instead, we estimated a population of likely models (each consistent with the training data)
using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells taken from Medici \emph{et al.} \citep{Medici:2008fk}.
We used the Pareto Optimal Ensemble Technique (JuPOETs) multiobjective optimization framework in combination
with leave-one-out cross-validation to estimate an ensemble of model parameters \citep{Song:2010fk,JuPOETs-BioArXiv}.
Model parameter values were adjusted to minimize the residual between simulations and experimental measurements for the 11 objective functions (supplemental materials).
Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \citep{kohavi1995study}.
The 41 intracellular protein and mRNA data-sets used for identification were organized into 11 objective functions.
These 11 objective functions were then partitioned, where each partition contained ten training objectives and one validation objective.
The training and validation data were Western blots. We achieved a biologically realistic concentration scale by establishing characteristic rates of transcription, translation, mRNA and protein degradation, as well as characteristic concentrations of ribosomes and RNAPs using the Bionumbers database \citep{Milo:2010aa}, (supplemental materials).
The overall concentration scale was nM, with proteins ranging from 10-1000nM and mRNA ranging from 0.01 to 1nM, reflecting the true abundances and ratios between each species.
An initial nominal parameter set was established by inspection. JuPOETs was then allowed to search in a neighborhood of $\pm$30\% of this nominal set.
The parameter ensemble estimated by JuPOETs is available with the model source code.
JuPOETs is open source and freely available for download under an MIT software license from http://www.varnerlab.org.

@@

\clearpage
\input{Supplemental-JB}

>>>>

\begin{titlepage}
{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\vspace{0.05in}
{\par \centering Russell Gould$^{\ddag}$ and David M. Bassen, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
\vspace{0.05in}
{\par \centering Robert Frederick Smith School of Chemical and Biomolecular Engineering and $^{\ddag}$Nancy E. and Peter C. Meinig School of Biomedical Engineering, Cornell University, Ithaca NY 14853}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
\vspace{0.1in}
% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
%\vspace{0.1in}
% {\par \centering \textbf{Character count:}~\emph{56,659}}
% \vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
\vspace{0.1in}
{\par \centering Jonathan T. Butcher,}
{\par \centering Associate Professor, Nancy E. and Peter C. Meinig School of Biomedical Engineering,}
{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jtb47@cornell.edu}
{\par \centering Phone: (607) 255 - 3575}
{\par \centering Fax: (607) 255 - 7330}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
We used ordinary differential equations (ODEs) to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was enhanced following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

\subsection*{Formulation, solution and analysis of the EMT model equations}
\subsubsection*{EMT signaling events.}
EMT signaling events were modeled using either saturation or mass-action kinetics within an ordinary differential equation (ODE) framework:
\begin{equation}
	\frac{1}{\tau_{i}}\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)-\mu x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}
\end{equation}
where $\mathcal{R}$ denotes the number of signaling reactions and $\mathcal{M}$ denotes the number of signaling proteins in the model.
The quantity $\tau_{i}$ denotes a time scale parameter for species $i$ which captures un-modeled effects; in the current study $\tau_{i} = 1$ for all species.
The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$.
Typically, reaction $j$ is a non-linear function of biochemical and enzyme species abundance, as well as unknown model parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$.
If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$.
Conversely, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
Species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.

@@

\subsubsection*{EMT gene expression processes.}
The EMT model described both signal transduction and gene expression events following the addition of TGF$-\beta$ and VEGF-A.
For each gene of the $\mathcal{G}=23$ we considered, we modeled both the resulting mRNA ($m_{j}$) and protein ($p_{j}$):
\begin{eqnarray}
	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\delta_{m,j}\right)m_{j}+\lambda_{j}\\
	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\delta_{p,j}\right)p_{j}
\end{eqnarray}where $j=1,2,\dots,\mathcal{G}$.
The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
the terms $\delta_{m,j}$ and $\delta_{p,j}$ denote degradation constants for mRNA and protein, respectively.
Lastly, $\mu$ denotes the specific growth rate, and $\lambda_{j}$ denotes the constitutive rate of gene expression for gene $j$.
The specific transcription rate was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
abundance of transcription factors, or other regulators influenced the expression of gene $j$.
The kinetic rate of transcription was modeled as:
\begin{equation}
	\bar{r}_{T,j} = \alpha_{j}\left[V^{max}_{T}\left(\frac{G_{j}}{K_{T}+G_{j}}\right)\right]
\end{equation}where the maximum gene expression rate was defined as the product of a characteristic transcription rate constant ($k_{T}$)
and the abundance of RNA polymerase, $V^{max}_{T} = k_{T}\left(RNAP\right)$. The parameter $\alpha_{j}$ denotes the gene specific correction to the characteristic transcription rate.
(estimated in this study), while $k_{T}$, $G_{j}$ and $RNAP$ were estimated from literature \citep{Milo:2010aa}.
Similar to the signaling processes, the gene expression control term $0\leq u_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
For each rate, we used a rule-based approach to select from competing control factors.
If the expression of gene $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$u_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on the expression of gene $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \citep{pr3010178,pr3010138}.
If a gene expression process has no modifying factors, $u_{j}=1$.

Lastly, the specific translation rate was modeled as:
\begin{equation}
	r_{X,j} = \beta_{j}\left[V^{max}_{X}\left(\frac{m_{j}}{K_{X}+m_{j}}\right)\right]
\end{equation}where $V^{max}_{X}$ denotes a characteristic maximum translation rate estimated from literature, $\beta_{j}$ denotes the transcript specific correction the characteristic translation rate,
and $K_{X}$ denotes a translation saturation constant.
The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
and the abundance of Ribosomes ($RIBO$), $V^{max}_{X} = k_{X}\left(RIBO\right)$, where both $k_{X}$ and $RIBO$ abundance were estimated from literature \citep{Milo:2010aa}.

@@

\subsubsection*{Estimation of model parameters using multiobjective optimization.}
We estimated a population of likely EMT model parameter sets (each consistent with the training data)
using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells taken from Medici \emph{et al.} \citep{Medici:2008fk}.
We used the Pareto Optimal Ensemble Technique (JuPOETs) multiobjective optimization framework in combination
with leave-one-out cross-validation to estimate an ensemble of model parameters \citep{Song:2010fk,JuPOETs-BioArXiv}.
Model parameter values were adjusted to minimize the residual between simulations and experimental measurements for the 11 objective functions (supplemental materials).
Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \citep{kohavi1995study}.
The 41 intracellular protein and mRNA data-sets used for identification were organized into 11 objective functions.
These 11 objective functions were then partitioned, where each partition contained ten training objectives and one validation objective.
The training and validation data were Western blots. We achieved a biologically realistic concentration scale by establishing characteristic rates of transcription, translation, mRNA and protein degradation, as well as characteristic concentrations of ribosomes and RNAPs using the Bionumbers database \citep{Milo:2010aa}, (supplemental materials).
The overall concentration scale was nM, with proteins ranging from 10-1000nM and mRNA ranging from 0.01 to 1nM, reflecting the true abundances and ratios between each species.
An initial nominal parameter set was established by inspection. JuPOETs was then allowed to search in a neighborhood of $\pm$30\% of this nominal set.
The parameter ensemble estimated by JuPOETs is available with the model source code.
JuPOETs is open source and freely available for download under an MIT software license from http://www.varnerlab.org.

@@

\clearpage
\input{Supplemental-JB-v7}
